\bf Estimating the number of clusters using Cross Validation

### 2015-09-27 technometrics.tex
add technometrics journal version

\documentclass[11pt]{article}
\usepackage{epsf}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{footnote}
\usepackage[bottom]{footmisc}
\usepackage[singlelinecheck=false]{caption}
\usepackage{caption}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{graphics}
\usepackage{float}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{lscape}
\usepackage{array}
\textheight=8.7in
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{longtable}

\def\threedigits#1{%
  \ifnum#1<100 0\fi
  \ifnum#1<10 0\fi
  \number#1}

\topmargin=0.1in \oddsidemargin=-0.1cm \evensidemargin=-0.1cm

\paperheight=11in \paperwidth=8.5in \marginparwidth=0in

\marginparsep=0in \textwidth=6.5in \headheight=0in \headsep=0in

@@

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\corr}{corr}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}
\newcommand{\T}{T}

@@

% class labels
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}

\begin{document}
\title{Estimating the number of clusters using Cross Validation}
\author{Wei Fu \qquad Patrick O. Perry \\\\ New York University}
\date{}
\maketitle
\begin{abstract}
Many clustering methods, including $k$-means, require the user to specify the
number of clusters as an input parameter. A variety of methods have been
devised to choose the number of clusters automatically, but they often rely on
strong modeling assumptions. We propose a data-driven approach to estimate the
number of clusters based on a novel form of cross-validation. This differs
from ordinary cross-validation, because clustering is fundamentally an
unsupervised learning problem. Simulation and real data analysis results show
that our proposed method outperforms existing methods, especially in
high-dimensional settings with heavy-tailed data.
\end{abstract}

\section{Introduction}

@@

\section{Cross-validation for selecting the number of clusters}

Cross-validation is commonly used for model selection in supervised learning
problems.  In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

\subsection{Gabriel CV algorithm}
\label{sec:gabriel-cv-algorithm}

We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
cross-validation, we permute the rows and columns of the data matrix and then
partition the rows and columns as $N = n + m$ and $P = p + q$ for 
non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
columns as ``predictors'' and the last $q$ columns as ``responses'';
similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \R^{n \times p}$,
$\Ytrain \in \R^{n \times q}$,
$\Xtest \in  \R^{m \times p}$,
and
$\Ytest \in  \R^{m \times q}$.

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
    be the assigned cluster for row $i$.

  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.

  \item \label{step:gabriel-predict}
    \textbf{Predict:}
    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
    $\bmuY(g) = \bmuY_g$.

  \item \label{step:gabriel-evaluate}
    \textbf{Evaluate:}
    Compute the cross-validation error
    \[
      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
    \]
    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
\end{enumerate}
\noindent
In principle, we could use any clustering and classification methods in
steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
report, we use $k$-means as the clustering algorithm.  For the classification
step, we compute the mean value of $X$ for each class; we assign an
observation to class $g$ if that class has the closest mean (randomly breaking
ties between classes).  The classification step is equivalent to linear
discriminant analysis with equal class priors and identity noise covariance
matrix.

To choose the folds, we randomly partition the rows and columns into $K$ and
$L$ subsets, respectively.  Each fold is indexed by a pair $(r,s)$ of
integers, with $r \in \{1, \dotsc, K\}$ and $s \in \{1, \dotsc, L\}$.  Fold
$(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
as ``response''.  We typically take $K = 5$ and $L = 2$.  For the number of
clusters, we select the value of $k$ that minimizes the average of $\CV(k)$
over all $K \times L$ folds (choosing the smallest value of $k$ in the event
of a tie).

\section{Self-Consistency of Gabriel CV method}

@@

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.

  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.

  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.

  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.

  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.

\end{enumerate}
Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
Gabriel CV correctly chooses $k = K$.
\end{theorem}

@@

Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
\end{proof}

\section{Analysis of Gabriel Cross-Validation with Gaussian Noise}

\subsection{Single cluster in two dimensions}

Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
the case of Gaussian noise. Our main result is that with single-cluster
Gaussian data, if the predictor and response columns of $\dataX$ are weakly
correlated or independent, then the method will correctly prefer $k = 1$ to
$k = 2$ clusters. We state the result in the case where $\dataX$ has 2 columns,
but it is straightforward to generalize this result to higher dimensions.

\begin{proposition}\label{prop:1clust-2dim}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}

Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to $\{ Y_i
\}_{i=1}^{n}$. With $k = 1$, the single-cluster centroid will
be equal to $\bar Y_n$, the sample mean of the $Y_1, \dotsc, Y_n$, approximately equal to
$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
will be
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]

@@

\begin{figure}[H]
\centering
\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$|\rho| < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

\subsection{Single cluster in higher dimensions}

To generalize the result of the previous section to higher dimensions, 
suppose that $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ is data from a single fold of
Gabriel Cross-Validation, with each $(X, Y)$ pair in $\R^p \times \R^q$ a
draw a mean-zero multivariate normal distribution. To analyze this setting, it
suffices to consider the mean-zero case with $\cov(X) = I_p$.

Suppose that $\cov(Y) = U \Lambda U^\T$ is the eigendecomposition of
$\cov(Y)$. For simplicity, suppose that the leading eigenvalue $\lambda_1$
has unit multiplicity, so that the first principal subspace of $Y$ is a line,
given by the span of the vector $u_1 \in \R^q$. When we apply $k$-means to
$Y_1, \dotsc, Y_n$ with $k = 2$, the centroids will be close to
$\bmuY_1 =  \sqrt{2\lambda_1 /\pi} u_1$ and
$\bmuY_2 = -\sqrt{2\lambda_1 /\pi} u_1$, with errors of size $\OhP(n^{-1/2})$.

The means for the $X$ variables will be close to
$\bmuX_1 = \E(X \mid u_1^\T Y > 0)$ and
$\bmuX_2 = \E(X \mid u_1^\T Y < 0)$,
with errors of size $\OhP(n^{-1/2})$.  Let
\[
  \begin{bmatrix}
    I_p & \lambda_1^{1/2} \rho \\
    \lambda_1^{1/2} \rho^T & \lambda_1
  \end{bmatrix}
\]
be the covariance matrix of $(X, u_1^\T Y)$,
where $\rho = (\rho_1, \dotsc, \rho_p)$ is such that $\rho_i = \corr(X_i,
u_1^\T Y)$ for $i = 1, \dotsc, p$.
. Then,
\[
  \bmuX_1 = \sqrt{2 \lambda_1 / \pi} \rho,
  \qquad
  \bmuX_2 = -\sqrt{2 \lambda_1 / \pi} \rho;
\]
thus, the classification rule learned from $\{ (X_i, \hGY_i) \}_{i=1}^{n}$
will send observations to class $1$ or $2$ according to whether
$\rho^\T X + \OhP(n^{-1/2})$ is positive or negative.

We have reduced the analysis of cross-validation error to the two-dimensional
case. In particular, following the calculations from the proof of
Prop.~\ref{prop:1clust-2dim}, if $|\corr(\rho^\T X, u_1^\T Y)| < 0.5$, then with
probability tending to one as $m$ and $n$ increase, $\CV(1) < \CV(2)$.
The condition $|\corr(\rho^\T X, u_1^\T)| < 0.5$ reduces to
\[
  \sum_{i=1}^p \rho_i^2 < 0.25.
\]
A sufficient (but not necessary) condition to ensure that this holds is that
the magnitude of the first canonical correlation between the $X$ variables and
the $Y$ variables is below $0.5 / \sqrt{p}$.

\subsection{Two clusters}

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL 
%---------We added packages---------------------
%\usepackage{epsf}
\usepackage{hyperref}
%\usepackage{graphicx}
\usepackage{subfigure}
%\usepackage[margin=2.5cm]{geometry}
%\usepackage{amsmath}
\usepackage{footnote}
\usepackage[bottom]{footmisc}
\usepackage[singlelinecheck=false]{caption}
%\usepackage{caption}
\usepackage{amsthm}
\usepackage{enumitem}
%\usepackage{graphics}
\usepackage{float}
\usepackage{amssymb}
%\usepackage{natbib}
\usepackage{lscape}
\usepackage{array}
%\textheight=8.7in
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{longtable}

%\def\threedigits#1{%
%  \ifnum#1<100 0\fi
%  \ifnum#1<10 0\fi
%  \number#1}

%\topmargin=0.1in \oddsidemargin=-0.1cm \evensidemargin=-0.1cm

%\paperheight=11in \paperwidth=8.5in \marginparwidth=0in

%\marginparsep=0in \textwidth=6.5in \headheight=0in \headsep=0in

@@

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}

@@

% class labels
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
%----------------------We added packages----------------------------
\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Estimating the number of clusters using Cross Validation}
  \author{Wei Fu \thanks{
    The authors gratefully acknowledge}\hspace{.2cm}\\
    Department of IOMS, New York University\\
    and \\
    Patrick O. Perry \\
    Department of IOMS, New York University}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Estimating the number of clusters using Cross Validation}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Many clustering methods, including $k$-means, require the user to specify the
number of clusters as an input parameter. A variety of methods have been
devised to choose the number of clusters automatically, but they often rely on
strong modeling assumptions. We propose a data-driven approach to estimate the
number of clusters based on a novel form of cross-validation. This differs
from ordinary cross-validation, because clustering is fundamentally an
unsupervised learning problem. Simulation and real data analysis results show
that our proposed method outperforms existing methods, especially in
high-dimensional settings with heavy-tailed data.
\end{abstract}

\noindent%
{\it Keywords:}  clustering, unsupervised learning, data-driven method 
\vfill
\hfill {\tiny technometrics tex template (do not remove)}

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

@@

\section{Cross-validation for selecting the number of clusters}
\label{sec:meth}
Cross-validation is commonly used for model selection in supervised learning
problems.  In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

\subsection{Gabriel CV algorithm}
\label{sec:gabriel-cv-algorithm}
We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
cross-validation, we permute the rows and columns of the data matrix and then
partition the rows and columns as $N = n + m$ and $P = p + q$ for 
non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
columns as ``predictors'' and the last $q$ columns as ``responses'';
similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \R^{n \times p}$,
$\Ytrain \in \R^{n \times q}$,
$\Xtest \in  \R^{m \times p}$,
and
$\Ytest \in  \R^{m \times q}$.

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
    be the assigned cluster for row $i$.
  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.
  \item \label{step:gabriel-predict}
    \textbf{Predict:}
    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
    $\bmuY(g) = \bmuY_g$.
  \item \label{step:gabriel-evaluate}
    \textbf{Evaluate:}
    Compute the cross-validation error
    \[
      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
    \]
    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
\end{enumerate}
\noindent
In principle, we could use any clustering and classification methods in
steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
report, we use $k$-means as the clustering algorithm.  For the classification
step, we compute the mean value of $X$ for each class; we assign an
observation to class $g$ if that class has the closest mean (randomly breaking
ties between classes).  The classification step is equivalent to linear
discriminant analysis with equal class priors and identity noise covariance
matrix.

To choose the folds, we randomly partition the rows and columns into $K$ and
$L$ subsets, respectively.  Each fold is indexed by a pair $(r,s)$ of
integers, with $r \in \{1, \dotsc, K\}$ and $s \in \{1, \dotsc, L\}$.  Fold
$(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
as ``response''.  We typically take $K = 5$ and $L = 2$.  For the number of
clusters, we select the value of $k$ that minimizes the average of $\CV(k)$
over all $K \times L$ folds (choosing the smallest value of $k$ in the event
of a tie).
%------------------------------------------------------------------
\section{Self-Consistency of Gabriel CV method}

@@

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.
  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.
  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.
\end{enumerate}
Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
Gabriel CV correctly chooses $k = K$.
\end{theorem}

@@

Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
\end{proof}
%-------------------------------------------------------------
\section{Analysis of Gabriel Cross-Validation with Gaussian Noise}

\subsection{Single Cluster}
Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
the case of Gaussian noise. Our main result is that with single-cluster
Gaussian data, if the predictor and response columns of $\dataX$ are weakly
correlated or independent, then the method will correctly prefer $k = 1$ to
$k = 2$ clusters. We first state the result in the case where $\dataX$ has two 
columns, and later generalize this result to higher dimensions.

\begin{proposition}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 0.5$,
then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}
Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to $\{ Y_i
\}_{i=1}^{n}$. With $k = 1$, the single-cluster centroid will
be equal to $\bar Y_n$, the sample mean of the $Y_1, \dotsc, Y_n$, approximately equal to
$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
will be
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]

@@

\begin{figure}[H]
\centering
\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

\begin{proposition}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^{p+q}$ is an
independent draw from a mean-zero multivariate normal distribution with 
covariance matrix $\Sigma_{XY}= \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
 \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If  $\frac{\sqrt{\lambda_1}}{2}
 > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}}$,
then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}
Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma_{XY}$, i.e.
\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma_{XY}\right)	\]
where $\Sigma_{XY}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.

Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$, with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. Then the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first PC 
of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
 where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.

To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$. Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
The conditional distribution $X|u^T_1 Y$ is hence normal with mean
 $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
Therefore, 
\begin{align}
\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
 			  &= \sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1
\end{align}
Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1$.
The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
and covariance matrix
\[
\begin{bmatrix}
\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
\end{bmatrix}
\]
The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
\[	
\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
\]
The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
\begin{align}
\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
\end{align}
Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
\[
E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
\]
Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
\begin{align*}
\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
\end{align*}

Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) > CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and  $\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$. By variance and bias decomposition of prediction MSE, when variance is the same, only bias influence the MSE.
